Self Regulation Ontology Retest Data: Initial Exploration

### 2017-02-23 SRO_Retest_Analyses.Rmd
cont working on retest exploration

---
title: 'Self Regulation Ontology Retest Data: Initial Exploration'
output:
  html_notebook: default
  pdf_document: default
---

@@

@@

Get list of test retest correlations

TODO: write a nicer of getting rid of non numeric columns

@@

The smaller meaningful vars have been somewhat cleaned. So far in this notebook we looked at all the vars in their rawest form.

@@

Apply the same cleaning procedure that was used for meaningful vars for all vars:
```{r}
transform_remove_skew = function(data, threshold = 1){
}
```

Transform non normal vars
Transform both retest and test data if not normally distributed in either. 
Order of transformations tried: log and sqrt
```{r}
#Add columns to dist_data that will be updated
dist_data = dist_data %>%
  mutate(retest_transform = ifelse(retest_not_normal, NA, 'none'),
         test_transform = ifelse(test_not_normal, NA, 'none'))

get_transform <- function(data){
  t_data = data
  transform = 'none'
  if(diff_from_normal(data) & !is.na(diff_from_normal(data))){
    t_data = log1p(data)
    transform = 'log1p'
    if(diff_from_normal(t_data) & !is.na(diff_from_normal(t_data))){
      t_data = sqrt(data)
      transform = 'sqrt'
      if(diff_from_normal(t_data) & !is.na(diff_from_normal(t_data))){
        transform = 'failed'
      }
    }
  }
  return(transform)
}

for(i in 1:nrow(dist_data)){
  if(is.na(dist_data$retest_transform[i])){
    dist_data$retest_transform[i] = get_transform(retest_data[,dist_data$dv[i]])
  }
  if(is.na(dist_data$test_transform[i])){
    dist_data$test_transform[i] = get_transform(retest_subs_test_data[,dist_data$dv[i]])
  }
}

t_retest_data = retest_data
t_retest_subs_test_data = retest_subs_test_data

#check if normal
#if yes stop and record transformation = none
#if transform: log
#check if normal
#if yes stop and record transformation = log
#if no transform: sqrt
```

>>>>

---
title: 'Self Regulation Ontology Retest Data: Initial Exploration'
output:
  html_notebook: default
  pdf_document: default
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

@@

# Loading and matching test and retest datasets

@@

# Correlations of exhaustive variables

Get list of test retest correlations for all matching (numeric) columns in the test and retest data. Note two caveats:
1. There are still some columns that weren't in both datasets (e.g. two stage)
2. Unlike meaningful variables these variables are not (yet) rid of outliers or log transformed if skewed. This will be detailed below along with its problems.

@@

The smaller meaningful vars dataframe have been somewhat cleaned. So far in this notebook we looked at all the vars in their rawest form. Here we'll first check the distributions of all the variables and then apply the same cleaning procedure that was used for the meaningful variables to all of them.

Note that I'm doing this for the sake of consistency for now. We might later decide that we would like to transform certain variables of interest in other ways to be able to include them in our analyses without violating the assumptions of those analyses. The current procedure checks the skewness of a column

@@

Apply the same cleaning procedure that was used for meaningful vars for all vars:
```{r}
remove_outliers = function(data_column, quantile_range = 2.5){

  q_25 = quantile(data_column, na.rm=T)[2]
  q_50 = quantile(data_column, na.rm=T)[3]
  q_75 = quantile(data_column, na.rm=T)[4]

  lowlimit = q_50 - quantile_range*(q_75 - q_25)
  highlimit = q_50 + quantile_range*(q_75 - q_25)

  data_column = ifelse(data_column<lowlimit, NA, ifelse(data_column>highlimit, NA, data_column))

  return(data_column)
}

"%w/o%" <- function(x, y) x[!x %in% y]

neg_log <- function(column){
  col_max = max(column, na.rm=T)
  column = col_max+1-column
  return(log(column))
}

transform_remove_skew = function(data, columns, threshold = 1){

  tmp = as.data.frame(apply(data[,columns],2,skew))
  names(tmp) = c("skew")
  tmp$dv = row.names(tmp)
  tmp = tmp %>% 
    filter(abs(skew)>threshold)

  skewed_variables = tmp$dv
  skew_subset = data[, skewed_variables]
  positive_subset = data[,tmp$dv[tmp$skew>0]]
  negative_subset = data[,tmp$dv[tmp$skew<0]]

  # transform variables
  # log transform for positive skew
  positive_subset = log(positive_subset)
  successful_transforms = as.data.frame(apply(positive_subset, 2, skew))
  names(successful_transforms) = c('skew')
  successful_transforms$dv = row.names(successful_transforms)
  successful_transforms = successful_transforms %>% filter(abs(skew)<threshold)
  successful_transforms = successful_transforms$dv
  successful_transforms = positive_subset[,successful_transforms]
  dropped_vars = names(positive_subset) %w/o% names(successful_transforms)

  # replace transformed variables
  data = data[,-c(which(names(data) %in% names(positive_subset)))]
  names(successful_transforms) = paste0(names(successful_transforms), '.logTr')
  print.noquote(rep('*', 40))
  print(paste0('Dropping ', length(dropped_vars) ,' positively skewed data that could not be transformed successfully:'))
  print(dropped_vars)
  print.noquote(rep('*', 40))
  data = cbind(data, successful_transforms)

  # reflected log transform for negative skew      
  negative_subset = as.data.frame(apply(negative_subset, 2, neg_log))
  successful_transforms = as.data.frame(apply(negative_subset, 2, skew))
  names(successful_transforms) = c('skew')
  successful_transforms$dv = row.names(successful_transforms)
  successful_transforms = successful_transforms %>% filter(abs(skew)<1)
  successful_transforms = successful_transforms$dv
  successful_transforms = negative_subset[,successful_transforms]
  dropped_vars = names(negative_subset) %w/o% names(successful_transforms)

  # replace transformed variables
  data = data[,-c(which(names(data) %in% names(negative_subset)))]
  names(successful_transforms) = paste0(names(successful_transforms), '.ReflogTr')
  print.noquote(rep('*', 40))
  print(paste0('Dropping ', length(dropped_vars) ,' negatively skewed data that could not be transformed successfully:'))
  print(dropped_vars)
  print.noquote(rep('*', 40))
  data = cbind(data, successful_transforms)

  return(data)
}
```

```{r}
clean_retest_data = as.data.frame(apply(retest_data[, matching_dv_columns], 2, remove_outliers))
clean_retest_data = transform_remove_skew(clean_retest_data, matching_dv_columns)
```

I'm not sure what is the right thing to do regarding the cleaning of the test data for the retest subjects. Should they be cleaned on the whole sample for the time they were collected on and then extracted (this should have an effect on outliers as well as transformations) or extracted and then cleaned. Here I am doing the latter but this seems to be resulting in more skewed variables.

```{r}
clean_retest_subs_test_data = as.data.frame(apply(retest_subs_test_data[, matching_dv_columns], 2, remove_outliers))
clean_retest_subs_test_data = transform_remove_skew(retest_subs_test_data, matching_dv_columns)
```

But what about dropping different variables from the two dataframes
